---
title: "CSE 270 Sports Analytics"
subtitle: "Homework 2"
author: "Name: Anna Martirosyan"
date: "19.10.2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(SportsAnalytics270)
library(dplyr)
library(ggplot2)
library(elo)
library(VGAM)
library(BradleyTerry2)
```

# Poisson Regression

## Problem 1(20 points)

### Predicting football games using Poisson regression.

1. Make predictions for the same game from the same league assigned to you in 
homework 1, but now use the poisson regression model instead of distributions.
**(8 points)**

```{r, echo=TRUE}
data("f_data_sm")
df = f_data_sm %>% mutate(FTAG = ifelse(FTAG>=4, "4+", FTAG),
                           FTHG = ifelse(FTHG >=4, "4+", FTHG))
table(Away=df$FTAG, Home = df$FTHG)

df %>% filter(SEASON==2022) %>% group_by(COUNTRY) %>%
  summarise(pvalue=chisq.test(table(FTAG, FTHG),
                              simulate.p.value=T)$p.value)
 
turkey = f_data_sm %>%
            filter(COUNTRY=="Turkey") %>%
            select(HOMETEAM, AWAYTEAM, FTHG, FTAG)

turk1=data.frame(turkey[,c("HOMETEAM", "AWAYTEAM", "FTHG")], Home=1)
turk2=data.frame(turkey[,c("AWAYTEAM", "HOMETEAM", "FTAG")], Home=0)
colnames(turk1)=c("Team", "Opponent", "Goal", "Home")
colnames(turk2)=c("Team", "Opponent","Goal", "Home")
turkey2=rbind(turk1, turk2)

model=glm(Goal~Team+Opponent+Home, data=turkey2, family=poisson(link=log))

options(scipen=1, digits=4)
erzurum_average <- predict(model, data.frame(Home=1, Team="Erzurum BB",
                                              Opponent="Denizlispor"), 
                            type="response")
options(scipen=1, digits=4)
denizlispor_average <- predict(model, data.frame(Home=0, Team="Denizlispor",
                                             Opponent="Erzurum BB"), 
                                type="response")

# After calculating the expected (average) number of goals for both teams 
# we can see that expected number of goals of Erzurum BB (1.431) is higher than 
# the expected (average) number of goals of Denizlispor (1.084). Therefore, 
# chances of Erzurum BB winning are higher.

options(scipen=1, digits=4)
intercept <- coefficients(model)["(Intercept)"]

team_erzurum <- coefficients(model)["TeamErzurum BB"]
oponnent_denizlispor <- coefficients(model)["OpponentDenizlispor"]
 
home_coeff <- coefficients(model)["Home"]

erzurum_h=exp(intercept + team_erzurum + oponnent_denizlispor + home_coeff)  
denizlispor_a=exp(intercept + team_erzurum + oponnent_denizlispor)
erzurum_h/denizlispor_a

###### As we have lambda for both teams, we can use skellam distribution to   ####### 
######               come up with the probabilities.                     #######

set.seed(1)
k = rskellam(10000,mu1 = erzurum_average, mu2 = denizlispor_average)

ggplot()+geom_bar(aes(x=k))+labs(x="Goal difference", 
     title="Skellam distribution for Erzurum BB vs Denizlispor")
 

(denizlispor_win = sum(dskellam(c(-100:-1),erzurum_average, denizlispor_average)))
(draw = sum(dskellam(0 ,erzurum_average, denizlispor_average)))
(erzurum_win = sum(dskellam(c(1:100),erzurum_average, denizlispor_average)))

```

2. Interpret the result and compare with the betting odds from homework 1.  
Is the same team expected to win ? **(2 points)**

```{r, echo = TRUE}
# The results that I got from Poisson Regression model and distribution approach
# are quite different. Using distribution approach I got that the expected 
# winner is Denizlispor with the following probabilities:

# Erzurum BB wins - 0.3245
# Denizlispor wins - 0.3706
# Draw - 0.3049

# But while using the Poisson regression approach I got the opposite result. In 
# this case instead of Denizlispor the expected winner is Erzurum BB with 
# following probabilities:

# Erzurum BB wins - 0.4498
# Denizlispor wins - 0.286
# Draw - 0.2642

```

3. Define a metric to calculate the efficiency of the predictions. 
**(5 points)**

```{r, echo = TRUE}
nfl_games_train =f_data_sm %>% filter(SEASON != "2022") 
nfl_games_test <- f_data_sm %>% filter(SEASON == "2022")

get_poisson_form <- function(data, country ){
  nfl_home <- data %>% 
    filter(COUNTRY == country) %>%
    group_by(HOMETEAM) %>%
    mutate(TEAM = as.factor(HOMETEAM),
           OPPONENT = as.factor(AWAYTEAM), 
           GOAL = FTHG,
           HOME = 1) %>%
    ungroup() %>%
    select(TEAM, OPPONENT, GOAL, HOME)
  nfl_away <- data %>% 
    filter(COUNTRY == country) %>%
    group_by(AWAYTEAM) %>%
    mutate(TEAM = as.factor(AWAYTEAM),
           OPPONENT = as.factor(HOMETEAM), 
           GOAL = FTAG,
           HOME = 0) %>%
    ungroup() %>%
    select(TEAM, OPPONENT, GOAL, HOME)
  return(rbind(nfl_home, nfl_away)) }

nfl_train_poisson <- get_poisson_form(nfl_games_train, "Turkey") 
nfl_test_poisson <- get_poisson_form(nfl_games_test, "Turkey")

nfl_model_poisson <- glm(GOAL~TEAM+OPPONENT+HOME, data=nfl_train_poisson, 
                         family=poisson(link=log))

nfl_model_poisson$xlevels[["TEAM"]] <- 
  union(nfl_model_poisson$xlevels[["TEAM"]], levels(nfl_test_poisson$TEAM))
nfl_model_poisson$xlevels[["OPPONENT"]] <- 
  union(nfl_model_poisson$xlevels[["OPPONENT"]], levels(nfl_test_poisson$TEAM))

nfl_test_poisson$P.G <- predict(nfl_model_poisson, 
                                nfl_test_poisson, 
                                type = "response")
predict(nfl_model_poisson, data.frame(TEAM = "Erzurum BB", 
                                     OPPONENT = "Denizlispor",
                                     HOME = 1, 
                                     type = "response"))

rmse = sqrt(mean((nfl_test_poisson$GOAL - nfl_test_poisson$P.G)^2))
rmse
# Lower values of RMSE indicate better fit
# RMSE for my model is 1.261
```

4. Find the under estimated and over estimated teams based on Poisson model. 
**(5 points)**

```{r, echo=TRUE}
underestimated_nfl_teams <- nfl_test_poisson %>%
  filter(GOAL < P.G)
underestimated_nfl_teams

overestimated_nfl_teams <-nfl_test_poisson %>%
  filter(GOAL > P.G)
overestimated_nfl_teams

```


# Bradley terry model

## Problem 2(30 points)

### Construct Bradley-Terry model for NBA regular season games.


1. Load the dataset **nba2009_2021** from the package **SportsAnalytics270** 
and remove the game Boston Celtics vs Indiana Pacers by filtering the 
**home.PTS** and **away.PTS** variables to be equal to 0.**(3 points)**

```{r, echo=TRUE}
data(nba2009_2021)
df_2 = nba2009_2021 %>% 
  filter(home.PTS !=0 & away.PTS!=0)
head(df_2, n=5)
```

2. Prepare the dataset for fitting a Bradley-Terry model by adding 2 variables,
**ht_w** and **at_w**.Assign the values of the variables to (**ht_w** = 1, 
**at_w** = 0) if the home team won the game and (**ht_w** = 0, **at_w** = 1) if 
the away team won the game. **(5 points)**

```{r, echo=TRUE}
df_3 = df_2 %>%
  select(home.TEAM_NAME, away.TEAM_NAME, home.WL) %>%
  mutate(ht_w = ifelse(home.WL=="W",1,0),
         at_w = ifelse(home.WL=="L",1,0))
head(df_3, n=5)

```

3. Convert the variables representing the home team and the away team into 
a factor. **(2 points)**

```{r, echo=TRUE}
df_3$home.TEAM_NAME <- as.factor(df_3$home.TEAM_NAME)
df_3$away.TEAM_NAME <- as.factor(df_3$away.TEAM_NAME)
str(df_3)
```

4. Fit a Bradley-Terry model into the dataset to get the abilities of the 
teams to win. **(5 points)**

```{r, echo=TRUE}
df_4 = df_3 %>%
    group_by(home.TEAM_NAME, away.TEAM_NAME) %>%
    summarise(ht=sum(ht_w), at=sum(at_w))
head(df_4, n=5)
```

5. Plot the abilities of the teams. **(5 points)**

```{r, echo=TRUE}
model = BTm(cbind(ht, at), 
            home.TEAM_NAME, 
            away.TEAM_NAME, 
            data=df_4, 
            id="team_")
coef=model$coefficients
sort(coef, decreasing = T)
BTabilities(model)
abilities = as.data.frame(BTabilities(model))
abilities$team = rownames(abilities)
abilities = abilities[order(abilities$ability, decreasing = T),]

ggplot(data=abilities, aes(x = reorder(team, ability), y = ability)) +
  geom_bar(stat="identity") + coord_flip() +
  labs(x="Teams", y="Ability to Win", title="Ability to win of NBA teams")
```

6. Interpret the results of the plot, which team is the best and which team 
is the worst ? Why is the ability to win for team **Atlanta Hawks** equal 
to 0 ? **(5 points)**


```{r, echo=TRUE}
# San Antonio Spurs: the strongest team 
# Sacramento Kings : the weakest team
# The ability to win for team Atlanta Hawks is equal to 0, because the results 
# were calculated with respect to that team.
```

7. Make a prediction for the upcoming 3 games.  
(Schedule can be found here: https://www.nba.com/schedule) **(5 points)**

```{r, echo=TRUE}
a_hawks=data.frame(home.TEAM_NAME=rep("Atlanta Hawks", 3), 
                   away.TEAM_NAME=c("Cleveland Cavaliers",
                               "New Orleans Pelicans",
                               "Detroit Pistons"))

a_hawks$home.TEAM_NAME=factor(a_hawks$home.TEAM_NAME, 
                              levels(df_4$away.TEAM_NAME))
a_hawks$away.TEAM_NAME=factor(a_hawks$away.TEAM_NAME, 
                              levels(df_4$away.TEAM_NAME))

a_hawks_prob = predict(model, newdata=a_hawks, level=2, type="response")

(a_hawks_df = data.frame(a_hawks, 
                         ht_w=a_hawks_prob,
                         at_w=1-a_hawks_prob))
```


# Winning Percentage

## Pythagorean wins for european football

## Problem 3 (30 points)

1. Library **SportsAnalytics270** has a function final_table. It creates final 
league standing for the season. Use this function to create a dataframe with
final standings of all seasons of your league. Combine seasons into 1 dataframe. 
You need to get something like nba_east. **(10 points)**

```{r, echo=TRUE}
final_df = data.frame()

for(i in unique(f_data_sm[f_data_sm$COUNTRY == "Turkey",]$SEASON)) {   
  output = final_table(f_data_sm, "Turkey", i)
  output$Season = i
  final_df = rbind(final_df, output)
}
head(final_df, n=5)

```

2. Create a variable for winning percentage and goal difference. Note, as 
there are draws in football, we are going to take draw as a half win. 
**(2 points)**
```{r, echo=TRUE}
final_df$WP = 
  as.numeric(format(round(((final_df$W + final_df$D/2)/(final_df$W+final_df$D+ 
                                               final_df$L))*100,1), nsmall = 1))
head(final_df, n=5)
# The column DIFF is already the difference of GF and GA
```


3. Plot Goal differential against Winning Percentage with a regression line.  
Make sure to have a title and appropriate axis labels for your plot.
**(5 points)**

```{r, echo=TRUE}

library(ggthemes)
ggplot(final_df, aes(x=DIFF, y=WP))+geom_point()+
  geom_smooth(method = "lm", se=F)+
  labs(x="Points differential",y="Winning percentage")+
  scale_x_continuous(breaks = seq(-70, 70, 10))+
  theme_stata() 

```

4. Interpret the plot, do you think there is a strong correlation between the 
variables ? **(3 points)**

```{r, echo=TRUE}
# Yes, it is nearly linear correlation between Goal differential and 
# Winning Percentage. If the points difference is near 0 the winning percentage 
# is near 50% which is quite expected, because if the difference is 0, means 
# the game ended up with home and away team scoring the same number of goals.
```

5. Calculate Pearson correlation coefficient for Goal differential and Winning 
Percentage. **(2 points)**

```{r, echo=TRUE}
(cor(final_df$WP, final_df$DIFF,method = "pearson"))
```


6. Build a regression model to estimate the value for k from the Pythagorean 
formula **(3 points)**

```{r,  echo=TRUE}
final_df_new = final_df %>% 
  filter(W!=0, L!=0, GA!=0) %>% 
  mutate(W = W + 0.5* T, L=W + 0.5* T, RATIO=GF/GA)

model2 <- lm(log(W/L)~0+log(RATIO), data=final_df_new)
summary(model2)

## The estimated value of K is 0.03539 approximately 0.04

```


5. Using the estimated value for k, calculate Pythagorean Winning percentage
and Pythagorean wins for each team. **(3 points)**

```{r, echo=T}
k = model2$coefficients[1]
final_df = final_df %>% 
       mutate(PWPCT = (GF ^ k / (GF^k + GA ^ k)), PW = (W + L + T) * PWPCT)
head(final_df, 5)
```


6. Find over performing and under performing teams. **(2 points)**

```{r, echo=TRUE}
#over_performing <- final_df %>%
#  filter(GOAL < P.G)
#under_performing <- final_df %>%
#  filter(GOAL > P.G)
```


# Elo Ratings
## Problem 4 (20 points)
### Building Elo rating model for NBA games.

1. Calculate the relative Elo ratings for all the games in the nba dataset.  
Make sure to adjust the ratings of the teams at the beginning of each season.  
This link https://fivethirtyeight.com/features/how-we-calculate-nba-elo-ratings/ 
shows the steps for calculating elo ratings in nba. **(10 points)**

```{r, echo=TRUE}
data(nba2009_2021)
home_court_advantage <- 100
k_nba <- 20
starting_elos <- 1505

calculate_elos <- function(data, home_factor, k, starting_elos){
  relative_elos <- data.frame()
  final_elos <- data.frame()
  for (season in unique(nba2009_2021$SEASON_ID)){
  seasonal_data <- data %>% 
    filter(SEASON_ID == season)
  season_elos <- elo.run(score(home.PTS, away.PTS) ~
          adjust(home.TEAM_NAME, home_court_advantage) + away.TEAM_NAME,
          data = seasonal_data, k = k,
          initial.elos = starting_elos
        )
  starting_elos <-0.75 * final.elos(season_elos) + (0.25 * 1505)
  relative_elos <- rbind(relative_elos,
                            data.frame(season_elos, SEASON_ID = season))
  final_elos_df <- data.frame(season = season, 
                              team_abr = names(starting_elos), 
                              elo = starting_elos)
  final_elos <- rbind(final_elos, final_elos_df)
 
  }
  return (list(final_elos=final_elos, relative_elos=relative_elos))
}

nba_relative_elos <- calculate_elos(nba2009_2021, home_court_advantage, k_nba, 
                                    starting_elos)$relative_elos
nba_final <- calculate_elos(nba2009_2021, home_court_advantage, k_nba, 
                            starting_elos)$final_elos
nba_relative_elos$Predicted <- ifelse(nba_relative_elos$p.A > 0.5, 1,0)
head(nba_relative_elos, n=5)
```

2. Interpret the predictive power of your model using either confusion matrix 
or brier score. **(5 points)**

```{r, echo=TRUE}
library(caret)
calculate_brier <- function(data){ 
  diff <- data$p.A - data$wins.A 
  accuracy <- sum(diffˆ2) / nrow(data) 
  return(accuracy)
}
(caret::confusionMatrix(data = factor(nba_relative_elos$Predicted), 
                        reference = factor(nba_relative_elos$wins.A)))
```

3. Make predictions for the same games you chose in **Problem 2.7**, but now 
use Elo rating model.  **(3 points)**

```{r, echo=TRUE}
teams = nba_final %>% 
  filter(team_abr == "Atlanta Hawks" & season=="2021")
teams
```

4. Compare the results **(2 points)**

```{r, echo=TRUE}
```

5. Try to play with the home advantage parameter, K value or anything else, 
either on football data or in nba and check if that will help to improve your 
models' predictive power.  **(Bonus question 10 points)**

```{r}
```
