---
title: "CSE 270 Sports Analytics"
subtitle: "Homework 3"
author: "Name: Anna Martirosyan"
date: "11.11.2022"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(SportsAnalytics270)
library(dplyr)
library(ggplot2)
library(stringr)
library(nnet)
```

# Competitive Balance

## Problem 1 (50 points)

### Noll-Scully number

1. Get the final tables for all countries in the dataset *f_data_sm* for all 
seasons(for each country). (10 points)

```{r, echo=TRUE}
final_df = data.frame()
data(f_data_sm)
for (j in  unique(f_data_sm$COUNTRY)){
  for(i in unique(f_data_sm[f_data_sm$COUNTRY == j,]$SEASON)) {   
    output = final_table(f_data_sm, j, i)
    output$Season = i
    output$Country = j
    final_df = rbind(final_df, output)
  }}

head(final_df, n=5)
```

2. Calculate the wining percentage(ratio) for all the teams in all seasons. 
(Consider draws as half wins).  (2 points)

```{r, echo=TRUE}
final_df$WP = as.numeric(format(
  round(((final_df$W + final_df$D/2)/(final_df$W+final_df$D+final_df$L))*100, 
                           1), nsmall = 1))
head(final_df,n=5)
```

3. Calculate the Noll-Scully number for each season and country using the 
winning ratio. (3 points)

```{r, echo=TRUE}
id_s = 0.5/sqrt(16)
final_df$WR = as.numeric(format(
  round(((final_df$W + final_df$D/2)/(final_df$W+final_df$D+final_df$L)), 
                           3), nsmall = 1))
final_dff<- final_df %>%
   group_by(Season, Country) %>%
   summarise(NS=sd(WR)/id_s)

head(final_dff, n=10)
```

4. Visualize the results grouped by seasons and countries. Make sure to include
meaningful title and axis names.  (You can use facets)  (10 points)

```{r, echo=TRUE}
ggplot(final_dff, aes(Season, NS))+
  geom_bar(stat = "identity")+
  facet_wrap(~Country,scales = "free")+
  ylab("Noll-Skully number")+
  ylim(0, 2)+
  scale_x_continuous(breaks = seq(1994, 2022, 9))+
  ggtitle("Noll-Skully number by SEASON and COUNTRY")

```

5.Interpret the visualization (5 points)
```{r, echo=TRUE}

ggplot(final_dff, aes(Season, NS))+
  geom_bar(stat = "identity")+
  facet_wrap(~Country,scales = "free")+
  ylab("Noll-Skully number")+
  ylim(0, 2)+
  scale_x_continuous(breaks = seq(1994, 2022, 9))+
  ggtitle("Noll-Skully number by SEASON and COUNTRY")+
  geom_hline(yintercept=1, col="red", size=0.5)

# To my interpretation I added also the below graph, with line passing through
# y=1 to clearly show that if the number is above 1.0 it means the teams are 
# further away in wins than we would expect given the ideal. If the number is 
# below 1.0, it means the  teams are closer in wins than we would expect.
# Higher is the number, lower is Competitive Balance. Closer is the number to 
# 1, higher is Competitive Balance.
# As we can see from the visualizations in our case for most Countries 
# Noll-Skully number is below 1, which means that the  teams are closer in wins 
# than we would expect.

```



### C5 competitive balance 

6.Calculate the C5 index for the number of goals scored by the top 5 teams in 
each country for each season. You should use the dataset from Problem 1.1. 
(5 points)

```{r,echo=TRUE}

top5 <- final_df %>%
  group_by(Season, Country) %>%
  filter(POSITION <6) %>%
  summarize(TopP=sum(GF))

all_teams <- final_df %>%
  group_by(Season, Country) %>%
  summarize(Pt=sum(GF))

C5<-data.frame(Season=top5$Season,Country = top5$Country, 
               C5=top5$TopP/all_teams$Pt)
head(C5,n=5)
```

7. Visualize the results for all countries and seasons. (Can be the same 
structure as in 1.4) (10 points)

```{r, echo=TRUE}
ggplot(C5, aes(Season, C5))+
  geom_line()+
  ylab("C5 index")+
  ylim(0.25, 0.6)+
  scale_x_continuous(breaks = seq(1994, 2022, 9))+
  ggtitle("C5 index by Season and Country")+
  facet_wrap(~Country)

```

8. Interpret the results (5 points)

The above graphs show the proportion of championships (or points) by top 5
teams, in other words the dominance of top 5 teams. For example from the graphs
we can conclude that in 2003 for Portugal the points gained from top5 teams 
were approximately 40% of total number of points earned by all teams. The 
highest numbers can be seen in Scotland in 2011 (~ 57%) and in Greece in 2002 
(~ 53%).

# Individual Performance (50 points)

## Problem 2

1. Load the dataset *nba_players* from the library *SportsAnalytics270* and 
filter it to only include seasons starting from 1992.  
Remove the columns blanl and blank2 and drop missing values or nas. (2 points)

```{r, echo=TRUE}
library(tidyr)
data(nba_players)
nba_players = nba_players %>% 
  filter(Year >= 1992)
nba_players <- nba_players[ , ! names(nba_players) %in% c("blanl", "blank2")]  
nba_players <- nba_players %>% drop_na()
head(nba_players, n=6)

```

2. Clean the variable Pos from players with double positions, for example 
C-PF should be turned into C.  (5 points)

```{r,echo=TRUE}
nba_players$Pos <- trimws(nba_players$Pos, whitespace = '-.*')
head(nba_players,n=6)
```


3. Convert the vaiable Pos into a factor and remove the columns 
X, Year, Player and Tm. (3 points)

```{r, echo=TRUE}
nba_players$Pos = as.factor(nba_players$Pos)
nba_players <- nba_players[ , ! names(nba_players) %in% 
                              c("X", "Year", "Player", "Tm")]
head(nba_players, n=5)
```


4. Divide the dataset into train and test datasets. 
(You can use dplyr or any other sampling method). (10 points)
```{r,echo=TRUE}
training_dataset  <- nba_players %>% dplyr::sample_frac(0.7)
testing_dataset   <- dplyr::anti_join(nba_players, training_dataset)

head(training_dataset, n=5)
head(testing_dataset, n=5)
```

5. Build a classifier model to predict the position of the player based on 
the playing attributes.  Use the training set for fitting the model. 
(20 points) You can use multinomial regression for that.
```{r, echo=TRUE}

training_dataset$Pos <- relevel(training_dataset$Pos, ref = "C")
multinom_model <- multinom(Pos ~ ., data = nba_players)
#summary(multinom_model)

(exp(coef(multinom_model)))
head(round(fitted(multinom_model), 2))
```


6. Predict the position of the players for the test dataset. (5 points)

```{r, echo=TRUE}
testing_dataset$ClassPredicted <- predict(multinom_model, 
                                          newdata = testing_dataset, "class")
# Building classification table
table <- table(testing_dataset$Pos, testing_dataset$ClassPredicted)
table

```

7. Interpret the predictive power of the model. (5 points)

```{r, echo=TRUE}
training_dataset$ClassPredicted <- predict(multinom_model, 
                                           newdata = training_dataset, "class")
# Building classification table
tab <- table(training_dataset$Pos, training_dataset$ClassPredicted)

# Calculating accuracy - sum of diagonal elements divided by total observations
round((sum(diag(tab))/sum(tab))*100,2)

# My model accuracy has turned out to be 60.7% in the training dataset, which is 
# good because as Professor Madoyan says, "if it is higher that 50%, than it is 
# a good approximation" :)


```