---
title: "HW4 - Network Visualizations"
author: "Name: Anna Martirosyan"
date: "Deadline 4 December 2022"
output: pdf_document
---

# Network Visualizations

## Problem 1 (70 points)

```{r, echo=F, include=F}
library(igraph)
library(intergraph)
library(network)
library(sna)
library(statnet)
library(statnetWeb)
library(circlize)
library(SportsAnalytics270)
library(ggplot2)
library(tidyr)
library(dplyr)
library(UserNetR)
```

1) Load the file "transfers.csv" into a dataset called **transfers**.  
Group by the variables **League_from** and **League_to**, then create new 
variables **Mean_Age**(Mean of the variable **Age**), **Mean_Fee** (Mean of 
the variable **Transfer_fee** divided by 1000000 ) and **Count**(number of 
transfers in that group) by using the summarize function and filter out the 
observations with less than 10 transfers in between (**Count** >= 10) 
(5 points)

```{r, echo=TRUE}
library(dplyr)
transfers = read.csv("transfers.csv")
transfers_1 = transfers %>%
   group_by(League_from, League_to)%>%
   summarise(Mean_Age=mean(Age), 
             Mean_Fee = (mean(Transfer_fee))/1000000,
             Count =n())%>% 
  filter(Count >= 10)  
head(transfers_1, n=5)
```

2) Create an igraph using `graph_from_edgelist` for the dataset you just got.  
(5 points)

```{r, echo=TRUE}
matrix = cbind(transfers_1$League_from, transfers_1$League_to)
graph = igraph::graph_from_edgelist(matrix,directed = T)
head(graph, n=5)
```

3) Set edge attributes for the summarized three variables and use their values. 
(5 points)

```{r, echo=TRUE}
graph = set_edge_attr(graph, "Mean_Age", value = transfers_1$Mean_Age )
graph = set_edge_attr(graph, "Mean_Fee", value = transfers_1$Mean_Fee)
graph = set_edge_attr(graph, "Count", value = transfers_1$Count)
```

4) Draw the network in a circle layout, pick one of the attributes as the width 
of the edge.  Were you able to get an interpretable graph ? If no what's the 
reason ? If yes interpret it :) (10 points)

```{r,echo=TRUE}
plot(graph, 
     layout = layout_in_circle(graph),
     lable = V(graph),
     mode="circle", 
     main="Visualization 1", 
     edge.width = E(graph)$"Count")

# Interpretation:
# The graph is impossible to interpret, because default edge width is 1, but 
# in out case, as we put the edge width to be equal to the "Count", the edges 
# become too thick which made impossible to understand their connections. In
# situations like this we need to transform and scale the variable for the edge. 
```

5) Apply a min-max or any other reasonable transformation to the variable 
chosen by you and visualize again using the scaled variable for now. !Hint, 
if you are using min/max replace 0 with a small number. (10 points)

```{r, echo=TRUE,fig.height=5, fig.width=5}
count = E(graph)$"Count"
count = (count-min(count))/(max(count)-min(count))
count[which(count ==0, arr.ind=T)] = 0.000001
graph = set_edge_attr(graph, "scaled.count", value=count)

plot(graph, 
     layout = layout_in_circle(graph),
     lable = V(graph),
     mode="circle", 
     main="Visualization 2", 
     edge.width = E(graph)$"scaled.count",
     edge.arrow.mode =0.6)
```

6) Interpret the graph (10 points)

You can write your interpretations directly here: Now, the graph is possible to
interpret because, we scaled the "Count" and got "scaled.count" for the 
edge.width. The leagues are vertices and edges show the count of transfers 
from one league to another. The width of the edge is directly proportional to
the count of transfers, which means if the count is higher, the edge is thicker,
and vice versa. The self loops are also possible, because some transfers happen
between 2 teams of the same league. The thickest edge is between Seria A and 
Seria A, because the number of transfers (288) is the highest between Seria A 
and Seria A. And the thinnest edge is between Premier Liga and 1.Bundesliga
because the number of transfers is only 10.


7) Using the dataset *transfers* create another subset of data. For now, pick 
any of the leageus that you want from the available ones and filter out the 
internal transfers in that league (meaning that the variables *League_from* 
and *League_to* are equal to each other). Now group by the variables 
*Team_from* and *Team_to* and summarize the same values as in Problem 1. 
(5 points)

```{r,echo=TRUE}
sub_data = transfers %>%
  filter(League_from=="Liga NOS", League_from == League_to)%>%
  group_by(Team_from, Team_to)%>%
  summarise(Mean_Age=mean(Age), 
            Mean_Fee = (mean(Transfer_fee))/1000000,
            Count =n())
head(sub_data, n=5)  
```

8) Create an igraph network and set the edge attributes for the graph, the 
same way as in problem 3. (5 points)

```{r, echo=TRUE}
matrix_2 = cbind(sub_data$Team_from, sub_data$Team_to)
graph_2 = igraph::graph_from_edgelist(matrix_2, directed = F)
graph_2 = set_edge_attr(graph_2, "Mean_Age", value = sub_data$Mean_Age )
graph_2 = set_edge_attr(graph_2, "Mean_Fee", value = sub_data$Mean_Fee)
graph_2 = set_edge_attr(graph_2, "Count", value = sub_data$Count)
head(graph_2, n=5)
```

9) Now use that graph and draw a chordDiagram and use any of the three 
attributes you desire, apply the transformations to the variables if needed. 
(10 points)

```{r, fig.width=20, fig.height= 20}
as_m <- as.matrix(intergraph::asNetwork(graph_2),
                  matrix.type = "adjacency", 
                  attrname="Count")
chordDiagram(as_m)
```

10) Interpret the visualization. (5 points)

The above graph shows connections between teams of Liga NOS. The data is 
arranged radially around a circle, and the count of transfers between the
vertices(teams) are drawn as asymmetric arcs connecting the vertices. The arcs 
show transfer counts between 2 teams. Since the count of transfers  between all 
teams are equal to each other and is equal to 1, only except from Braga to 
Benfica (count is equal to 2), only the arcs from Braga to Benfica ate thicker 
than others.

# Network Statistics (30 points)

## Problem 2

1) Using the network created in **Problem 1.2** calculate the edge density of 
the network. Interpret the result. (5 points)

```{r, echo=TRUE}
(density= edge_density(graph, loops=F))

# The density can range from 0 to 1 and shows how interconnected are the nodes 
# within the network. Self loops are counted in  the total  number of edges so 
# with self loops can have density higher than 1.
```

2) Now calculate and interpret the reciprocity of the network. (5 points)

```{r, echo=TRUE}
(reciprocity = reciprocity(graph))
```


3) Find the most active league of the market. by calculating the degree 
centrality measure (5 points)

```{r,echo=TRUE}
which.max(igraph::degree(graph, mode="out", normalized=T))

#LaLiga is the most active league of the market
```

4) Calculate the closeness statistics of the network, pick any of the edge 
attributes as a weight and interpret the results. (5 points)  

```{r, echo=TRUE,fig.height=10, fig.width=10}
igraph::closeness(graph)
plot(graph, displaylabels=T, edge.width = E(graph)$"scaled.count")

# Interpretation: Closeness statistics of my graph network show how close each 
# league is to everyone else. In out case, LaLiga and Premier League are the 
# closest to others since their closeness statistics is the highest (0.125), 
# and the Eredivisie is the farest, since it has the lowest score 
# (approximately 0.077).
```

5) Now calculate the page rank of the network, which league is the most 
popular ?. (5 points)

```{r, echo=TRUE}
page_rank(graph)$vector
# The most popular is LaLiga since its page rank is the biggest.
```


6) Detect communities in the network. (5 points)

```{r, echo=TRUE,fig.height=10, fig.width=10}
edge_b <- edge.betweenness.community(graph)
plot(edge_b, graph, main="Edge Betweenness")
```